Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-18 VegOzone.tex
Added comments regarding conclusions and discussion

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects of the bespoke parameterizations of the $DO_3SE$ model with respect to the MM default parameterization on ozone uptake:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake in both probed years,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the increase reported above,
\item an earlier assumed start of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher extent of stomatal opening at low light intensities $\Rightarrow$ increased/reduced uptake,
\item symmetric variation of the extent of stomatal opening at low light intensities $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is always larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, they suggest considerable underestimations of ozone risk for subarctic species when compared to generic MM parameterisations. In particular, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the risk of ozone damage for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

@@

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have studied the ozone seasonality for northern Fennoscandia and the quality of its representation in available global and regional reanalysis products for Europe. We found that both global reanalysis products of ECWMF fail to reproduce observed ozone paterns. Only the CAMSRAQ ensemble reproduces the observed ozone seasonality well, although with a slight low-bias remaining. This confirms that high spartial resosultion and assimilation of in situ observations of ground-level ozone concentrations are a must to constrain reanalysis related to air quality and risk assement ozone damage to vegetation, especially in high latitudes.

We have exemplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advent of climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was in particular caused by the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found start of growing season ($106/122\,\unit{doy}$) is within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

We found representative ozone climatologies (multi-annual means) for northern Fennoscandia and Svanhovd based on ground-level ozone monitoring. To account for missing data from our 2018 record, we proposed a routine for gap filling of ozone observations following a Reynolds's decomposition into anomalies and climatology. We also assessed the quality of available global reanalysis (ECMWF MACC and CAM) and regional model reanalysis (Copernicus) ground-level ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were, in general, not significantly elevated by statistical means. It therefore occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrics like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periods of time. Consequently, acute cellular damage caused by very high fluxes over a short period of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differentiate picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, was found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandia vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandia vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation ought to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need verification by experimental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed start of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher extent of stomatal opening at low light intensities $\Rightarrow$ increased/reduced uptake,
\item symmetric variation of the extent of stomatal opening at low light intensities $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite is be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch and perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier start of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start of the growing season. It is therefore not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier start of the growing season, will be reduced compared to present day. Hence, a simulated earlier start of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions were established based on central European species or PFTs and might as well be subject to biases.

At the same time, intra-species variability even at the same location is typically non negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in Norway spruce ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest decidous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resilience of northern Fennoscandia plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.\\

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature response functions, once varified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

>>>>

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects of the bespoke parameterizations of the $DO_3SE$ model with respect to the MM default parameterization on ozone uptake:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake in both probed years,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the increase reported above,
\item an earlier assumed start of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher extent of stomatal opening at low light intensities $\Rightarrow$ increased/reduced uptake,
\item symmetric variation of the extent of stomatal opening at low light intensities $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is always larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, they suggest considerable underestimations of ozone risk for subarctic species when compared to generic MM parameterizations. In particular, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the risk of ozone damage for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

@@

\conclusions[Discussion and conclusions]
\label{sec:conc}

We have studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key climate variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of the growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

We found representative ozone climatologies (multi-annual means) for northern Fennoscandia and Svanhovd based on ground-level ozone monitoring. To account for missing data from our 2018 record, we proposed a routine for gap filling of ozone observations following a Reynolds's decomposition into anomalies and climatology. In the course, we also assessed the quality of available global (MACC and CAMS) and regional (CAMS Regional Air Quality (CAMSRAQ) ensemble) reanalysis products for Europe concerning ozone. We have studied the ozone seasonality for northern Fennoscandia and the quality of its representation in available global and regional reanalysis products for Europe. We confirm that both global reanalysis products do not reproduce the observed ozone patterns in northern Fennoscandia \citep{GMD:Huijnen2020,ACPD:Barten2020}. Only the CAMSRAQ ensemble reproduces the observed ozone seasonality, although with a remaining low-bias. This confirms that high spatial resolution and assimilation of in situ observations of ground-level ozone concentrations are a must to constrain reanalysis products related to air quality and risk assessment ozone damage to vegetation, especially in high latitudes.

Ozone peak concentrations were observed more frequently in 2018 than in 2019. In 2018, we found significantly enhanced \chem{[O_3]} particularly in April and May, but not during the active wildfires in July. This suggests that weather conditions and high ozone concentration at the beginning of the growing season are more important in risk assessment using flux-based metrics than episodes of peak concentrations over shorter periods. Because these peak \chem{[O_3]} may coincident with environmental conditions that limit ozone uptake and hence do not lead to damage. Consequently, acute cellular damage caused by very high fluxes over a short period is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a different picture. For herbal and grassland species (similar for crops) a strong dependency on the phenological state, e.g. high susceptibility to ozone in the reproductive state, was found \citep{EP:Bassin2004}.

To estimate potential biases in the assessment of ozone-induced risk on Fennoscandia vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandia vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation ought to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need verification by experimental data.

In summary, our key findings are:
\begin{enumerate}
   \itemsep0pt
 \item the bespoke parameterizations for subarctic species are important (especially concerning light and temperature and the climate effect on the growing season) to be able to assess \chem{O_3} uptake,
 \item the parameterizations defined in the mapping manual for European regional risk assessment would appear to not adequately capture the physiological responses of subarctic physiology that are important determinants of \chem{[O_3]} sensitivity under prevailing environmental conditions and are likely to substantially underestimate risk,
 \item the ozone reanalysis products tend to underestimate \chem{[O_3]} in Fennoscandia, and
 \item that the climatic conditions promoting ozone damage (particularly extended heatwaves and the timing of the growing season concerning the ozone spring peak) and which are likely to become more frequent under climate change are likely to further enhance the sensitivity of subarctic vegetation to ozone.
\end{enumerate}

However, the decline of the ozone spring peak is partly caused by dry deposition of ozone on vegetation. It, therefore, remains unclear whether an earlier start of the growing season has the potential to reduce the ozone spring peak concentration compared to present-day and hence reduce the damage risk of ozone in the future. 
We have not assessed the effect of the plants' response to VPD here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care because the biomass reduction functions were not established based on subarctic species and might as well be subject to biases.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in Norway spruce ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest deciduous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one season of exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resilience of northern Fennoscandia plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due competing effects, as an ozone-induced sluggishness \citep{SR:Hoshika2015} renders stomata unable to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the MM version of the $\mathrm{DO_3SE}$ model does not account for any ozone-inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.\\

Beyond the risk assessment of ozone-induced damage, our hypothetical bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.
